draft 1

### 2016-06-17 finalwriteup/draft.rmd
reviewed lit review started method review

B2B - Business to Business 
CBR - Case based reasoning. A method of estimating by using the results of similar cases.
CRM - Customer Relationship Management. Refers to a common type of software used by businesses to record client and project details as well as employee timesheet records. Generally accessable to each employee in a company.
IT - Information technology
BIM - Building Information Modelling. A digital representation of physical characteristics of an object or structure.

@@

Advantages of CBR include algorithmic efficiency as well as user engagement. In @Kim2004's study, it was noted that CBR models were simple to update with new data in comparison to neural networks, which must be slowly re-trained and re-tuned. New cases also did not need to have every variable complete which can occur often in real, messy data. Neural networks on the other hand require complete data. Another significant advantage of CBR is the ability for users to review chosen similar cases and make sense of the predicted cost, as opposed to neural networks which are a black box [@Kim2004][@Elfaki2014][@Kumar2007]. A disadvantage of CBR is the accuracy can be highly dependent on the number of selected cases [@Elfaki2014]. Overall, CBR may provide prediction accuracies between regression and neural networks but its results can be justified by the user - a valuable asset.

CBR can be described as a systematic method of expert judgment, where the decision maker manually compares similar projects from his or her experience [@Shepperd1996]. CBR that is performed manually through personal experience, without the assistance of algorithms, was the second most popular method for construction estimation in @Akintoye2000's survey in the UK. This indicates it is quite an intuitive method for decision maker's to cross check their detailed cost estimate, although an algorithmic version of CBR has not been taken up by industry.

@@

In summary, research into cost estimation in the construction industry has been undertaken using a variety of methods ranging from detailed manual analysis, algorithmic models, and case based reasoning. Neural networks generally outperformed all other systematic methods, such as regression and case based reasoning, but had a significant disadvantage of being a black box. Alternative methods such as boosted trees have the potential to perform as well as neural networks and provide insight into the structure of the algorithm. This is true for CBR as well, although it has not been shown to predict as accurately as neural networks. There exists a gap in the research in ways to intelligently combine CBR with machine learning methods that could predict well, provide insight to the structure, and engage the user in reviewing similar projects. Also, as stated previously, there exists a gap in the application of cost estimating in the construction industry to smaller components of the project, such as consulting and contracting companies who each face a similar problem in estimating their fixed fee. These companies could mobilise information in their internal project databases, as opposed to available literature which surveys dozens of companies involved in selected construction projects.

@@

Like the construction industry, the software industry faces the challenging task of estimating the cost of a project before full engagement with their client. However, unlike construction projects, the major cost associated with the project is *effort* as opposed to the cost of building materials. A parallel can be drawn between software project effort estimation research and consultants effort in the construction industry, which was a gap in the research of the previous section.

Effort estimation is indeed a problem, and was demonstrated by a review of 6 surveys of IT projects between 1984 and 1994 by @Molokken2003. It revealed that 60-80% of IT projects encounter effort and/or schedule overruns, where the average overrun was by 30 to 40%. A survey of software managers asked whether they saw effort estimation as a problem, and 91% responded 'yes', while only 9% answered 'no' [@Moores1992]. The most popular response for the supposed reason for cost overruns was over-optimistic estimates (51%), which falls squarely within the theory of optimism bias, or the 'inside view' [@Flyvbjerg2011]. Given this widespread problem in the industry, plenty of research has been dedicated to the problem of effort prediction, with methods similar to the construction industry literature: detailsed analysis, algorithmic modelling, and case based reasoning.

@@

Similar to the construction industry, expert judgment or detailed analysis, is the most widely practised method for effort estimation [@Shepperd1996][@Molokken2003]. This is despite many years of research being dedicated to developing algorithmic models and industry tools that, in the research context, outperform expert judgment. A survey by @Moores1992 found that software tools assisting a software project were in significant use, but only the project planning tools, not the cost estimating tools. This suggests a mental link for project managers to price jobs based on *details* of the software project as opposed to taking an outside view. Another study by @Heemstra1992 found that there was no evidence estimation accuracy improved when estimation tools were used. @Bergeron1992 similarly found jobs that used algorithmic models were actually associated with less accurate estimtes. However, this trend may be coincidental due to a lack of cases where estimation tools were used [@Molokken2003]. The scarcity of real project evidence that algorithmic estimation tools improve estimations and an unintuitive mental jump could be reasons contributing to detailed expert analysis remaining the most widespread technique for effort estimation.

@@

Algorithmic methods to predict effort for IT project costs has been a popular subject of research. Regression models have been the most substantially tested followed by neural networks [@Finnie1997]. This section will assess the details and performance of the algorithmic models, important variables, applications and finally gaps in the literature.

Multiple studies have shown that neural networks definitively outperform regression models in effort estimation, although regression is the most popular method in the literature [@Finnie1997][@Pai2013][@Matson1993]. Interestingly, several studies have found that even if 15 or so variables are included, often only one variable contributes significantly to the models accuracy: size [@Shepperd1996][@Finnie1997][@Pai2013]. Size can refer to the expected number of lines of code or function point. Function point refers to the amount of business functionality expected from an information system product. This is easier to guess correctly at the beginning of a project, and therefore leads to more accurate predictions [@Finnie1997]. One study used a single variable approach (size) with a linear coefficient and an exponential coefficient as follows:

$effort = \alpha*size^{\beta}$
$\alpha = productivity coefficient$
$\beta = economies of scale coefficient$
$size = estimated lines of code$

This model was compared to case based reasoning, and case based reasoning outperformed it [@Shepperd1996].

@@

Algorithmic estimation tools have not been successfully adopted by industry. This may be because no model has proved to be outstandingly successful at consistently predicting required effort [@Finnie1997]. Also, neural networks (the most accurate algorithm) lack the ability to capably explain its results and instill trust in decision makers [@Finnie1997].

The literature heavily focuses on effort estimation for the software industry. This leaves a gap in studies for effort estimation for consulting companies in other industries such as the construction industry. Construction projects do not have a simple 'project size' or function point variable and would need to utilise other variables available.

@@

In the IT industry, applications of case based reasoning to effort estimation have been researched in a similar manner to the construction industry. The procedure is described as a algorithmicallyfinding similar projects to a new case using measurements of euclidean distance in n-dimensional space, where each dimension corresponds to a variable [@Shepperd1996]. The similar cases are then used to predict effort for the new case, and in @Shepperd1996's study, linear regression was used for this final step.

In contrast to the construction industry, the literature in software effort estimation found CBR to perform approximately equally to algorithmic methods such as neural networks [@Shepperd1996][@Finnie1997]. It was also highlighted that CBR is intuitively similar to how an expert thinks about pricing projects and provides insight into the prediction that neural networks do not [@Finnie1997]. Research into CBR in the IT industry found similar results to the construction industry except that CBR performed as well as algorithmic methods.

@@

Research into effort estimation in the software industry has followed a similar path to the cost prediction of construction projects. Expert judgment via detailed analysis is still the predominant method for effort estimation. Case based reasoning and algorithmic methods such as neural networks have found some success in effort estimation (at similar levels) but have not been successfully adopted by industry. It was noted that case based reasoning resonated more with decision makers because its effort prediction could be justified by the model. This visibility of the model prediction structure is very important for future work if the results are to be translated to industry. Effort estimation remains a problem, however agile methods of project management and effort management are changing the contractual approach of consulting in the software industry [@Badenfelt2011]. This innovative method keeps delivery times short (and therefore predictable effort) and defines projects as something with a fixed cost and time. This means if a change occurs, other less important elements must be discarded maintain the fixed amount. Software development is a relatively new field in comparison to the construction industry and is able to adopt this new model relatively quickly across the industry. However, effort estimation for consulting companies in the construction industry are faced with a long history of tradition that is much harder to tackle. This leaves a gap in research in algorithimic models that predict effort for consulting companies in the construction industry. From previous literature, the models should allow decision makers to interrogate the structure of the prediction and be based on internal company data.

## Statistical and machine learning methods

@@

The aim of this project is to use statistical techniques to model the profitability of projects. In order to optimise the model, several statistical and machine learning techniques will be tested, and compared and it is therefore prudent to review the range of statistical and machine learning techniques that have been successfully applied in the literature. These vary from simple methods such as Linear Regression to complex, deep learning such as Neural Networks. The previous section highlighted the use of linear regression, neural networks, SVM and in one case boosted trees for prediction, however other business problems have utilised a wider range of methods. These include Naiive Bayes, Random Forests, and machine learned Bayesian Networks. The following section will present each method's advantages and disadvatanges and an example of a past successful application.

@@

Linear regression is one of the simplest and most popular statistical prediction methods [@Hastie2009]. An equation for a line is learned, which is defined by a constant and a coefficient times the value of each variable. The coefficients are chosen to minimise the residual sum of squares (least squared) [@Hastie2009]. The reason this method is popular is that it is simple, provide stable predictions, and can be adapted for categorical variables [@Seng2010]. The integration of categorical variables is called Anova, and the algorithm compares group mean variances within the categorical variable.

There are several drawbacks to linear regression. For example assumptions must be made about the structure of the data. Predictor varialbes are assumed to be linearly related to the response variable and the varialbes are normally distributed. Many real world phenomena do not correspond to these assumptions which can make it difficult, sometimes impossible, to produce useful results [@Hastie2009][@Putler2012]. Furthermore, response variables must be continous numeric variables and if two or more predictor variables are correlated, the values of their coefficients are highly unstable. On the other hand, correlated variables do not impact the predictive accuracy of the model. A final drawback is that the variable coefficients cannot be used to compare variable importance unless the varialbes have all been standardised to one another [@Putler2012]. This limits the amount of insight that can be drawn from a linear regression model.

Despite the limitations of linear regression models, they have widespread business applications due to their simplicity and ease of application. Regression is used to predict risk in the finance and insurance industries, predict who to target in marketing exercises, and predict consumption spending in the field of economics [@Harrell2013]. Due to its simplicity and widespread application, linear regression is a good starting point for an analysis and benchmark for performance against complex models.

@@

The method behind linear regression can be adapted to predict a binary response variable (1 or 0). A predictive linear equation is computed so that the predictor variables are combined with coefficients to give an output between 0 and 1. This is then transformed into the log odds to mediate extreme variable input values. The result is a probability an input case will result in a 0 or 1 (i.e. success or failure) [@Moore1989]. Similar pros and cons exist for logistic regression as linear regression, and it too can be used as a benchmark to compare other binary predictive models due to its speed and simplicity. Many business problems have binary response variables, such as yes/no, male/female, buy/do not buy, success/failure, or survive/death [@Moore1989].

@@

Naive Bayes prediction method makes conditional independence assumptions about the predictor variables in order to greatly simplify probability calculations for the response variable (the response variable must be categorical). Specifically, it is assumed that each variable is conditionally independent given the respose class alone, without considering other predictor variables. Mathematically this means that each variable's probability contribution can be simply calculated by looking at the probability of a response class given a predictor value by itself. The predictor variables probability estimates are combined for each possible response class, and the response class with the highest probability is chosen [@Provost2013].

The advantages of this method are that the conditional independence assumption enables very fast calculations and predictions. The method can perform very well for real world tasks because the assumption of independence does not damage predictions significantly. This is because if multiple varialbes are related, the variables independently still direct the prediction in the correct direction. The correlated variables will double or triple their emphasis on the predicted response variable, however this does not necessarily impact classification accuracy as the final step is to simply choose the class with the *highest* probability [@Provost2013]. This is fine for ranking. On the other hand, the output probabilities are not realisitic, and the values themselves should not be used [@Caruana2006]. Another disadvantage is that in order to calculate the class probability from a numeric variable, a distribution must be assumed (often gaussian) which the numeric data may not fit into neatly [@Caruana2006]. Finally, Naive Bayes classifers are known to not perform well in binary classification [@Caruana2006].

A real world success story for Naive Bayes classifiers hinges on their ability to be refined incrementally with each piece of new data, as opposed to re-calibrating the entire model. Naive Bayes has been used in complex spam detection systems where new spam emails or toxic text themes can be quickly added to the filtering model. The Naive Bayes method provides a good benchmark to compare against more complex models that should outperform it [@Caruana2006].

@@

Decision trees are one of the simplest and most intuitive machine learning methods. There are several brands of basic decision tree algorithms including ID3, C4.5, CART, and CHAID with CART and C4.5 being the most popular [@Kabra2001]. Ravi Kumar and Ravi [@Kumar2007] recommend the CART algorithm as it is capable of solving both classification and regression problems whereas the remaining decision trees solve classification problems only. Single decision trees have the risk of over-fitting the data and low predictive accuracy but have favourable characteristics such as providing intuitive rules that a decision maker can follow in real-life scenarios. In addition their results do not depend heavily on the skill of the data analyst [@Kabra2001][@Putler2012].

To tackle the problem of low predictive accuracy and instability, ensemble tree methods were pioneered in the 1990's with success [@Breiman1996]. Three examples of these ensemble decision tree methods are bagging, boosting, and random forests. Bagging creates multiple trees by sampling a different training data set for each tree and then combining tree results [@Breiman1996]. Random forests is an advanced form of bagging in that multiple trees are sampled from the same training data. However, when creating the tree for each sample dataset, a random subset of attributes is used to determine each split. This prevents the trees from becoming too correlated [@Breiman2001]. The boosted decision tree approach combines gradient descent with a series of decision trees. Each tree is limited to a certain depth to maintain simplicity, and each tree models the residuals (or errors) from the preceding tree. The limited depth of each tree prevents overfitting at each stage and the combined result of up to thousands of trees is very powerful [@Elith2008].

Ensemble decision trees have a different foundation to more traditional statistical prediction methods such as regression. For this reason, there are some advantages and disadvantages. The progression of binary splits means non-linear feature interactions can be captured and the predictor variables do not need to be transformed as no assumptions are made about the data's statistical distributions [@Louppe2014][@Radenkovic2010]. Other advantages are that the ensemble methods are very fast in comparison to SVM's and neural networks but perform just as well and provide insights such as variable importance and variable relationships [@Sealfon2012]. Random forests in particular has the benefit of not being affected by noise in the data as well as the capacity to handle a large number of variables [@Sealfon2012]. Caruana [@Caruana2006] tested boosted trees, random forests, neural networks, SVM's, logistic regression and naive bayes on 11 binary classification problems and found that boosted trees performed best, followed by random forests. Clearly, ensemble tree methods can compete with high level machine learning algorithms.

Along with the many advantages of trees and ensemble trees, it is important to be aware of their limitations. As trees are not built on a probabilistic framework, results can not be provided in this framework. For example, confidence intervals for predictions are not available for standard ensemble methods [@Louppe2014]. Also, variable importance tables provided by random forests and boosted trees can be biased towards variables with many categories and another type of ensemble method, conditional forests, should be used [@Radenkovic2010]. The methods can also be prone to overfitting if not handled carefully. Ensemble trees have many advantages particularly predictive power, but it is important to recognise that the best method for a problem often depends on the type of problem [@Louppe2014][@Caruana2006].

@@

A bayesian network is a graphical probabilistic model that illustrates the conditional dependencies between variables in a data set. The model is visually represented by a DAG (directed acyclic graph) and is capable of linking the conditional dependency between any variable to another variable or via other variables which are represetned by arcs [@Heckerman1998]. This kind of conditional relationship is bayesian. Bayesian networks have found success in combining deterministic models with observation data as well as expert knowledge as the statistical relationships between variables can be determined from different places and manually entered into a single model [@Kragt2009].

Bayesian networks are excellent at representing knowledge from different sources and types, and visuallising them in a way that communicates results well to decision makers [@Kragt2009]. Therefore bayesian networks become an excellent decision support tool. Network relatitonships can be machine learned, however this is not widely included in the suite of machine learning methods and is referred to more as a way to model complex networks. Drawbacks of bayesian networks include their inability to handle continuous numeric variables (which must be discretised). This means that the varialbe relationships are categorical [@Kragt2009]. Popular applications of bayesian networks are modelling uncertainty in natural resource management and modelling complex business network structures such as airports [@Wu2013].

@@

Neural networks were inspired by biological neural networks of the human nervous system and require a lot of training data. They are very flexible in the relationships they can mimic and are not greatly dependent on the skills of the analyst (Putler & Krider, 2012). A disadvantage is that although the results are accurate, the algorithm is a black box - meaning the data is input and the results output without providing the user insight into the methodology. Neural networks can perform high level tasks as a result of deep learning such as hand writing recognition, vehicle control, face recognition, and cancer detection [@Haykin2004].

The simplest type of Support vector machines (SVM's) calculate the linear boundary between two categories in feature space so that the category of future cases may be determined depending on which side of the boundary they fall. The location of the boundary is determined by the position of the widest clear margin that can be drawn between the two categories, allowing a certain number of misclassifications. This concept can be adapted to non linear boundaries by performing kernel transformations on prior variables as well as regression problems by introducing loss functions [@Auria2008]. Advantages of SVM's include the ability to model non linear relationships in data, and that no assumptions about the distributions of the data must be made as the model is non-probabilistic. They are also known to be robust over different samples and perform well in high-dimensional feature space [@Auria2008]. On the other hand, the output is in the form of distance to the boundary (as opposed to probability) and results are not transparent (similar to neural networks) [@Caruana2006]. Also, SVM's are known to be very slow to train and therefore often not suitable for industry purposes [@Auria2008]. Because SVM's perform well in high-dimensional space and can form non-linear rules they have been applied to protein classification (medicine) and text and image recognition [@Byun2002].

@@

@@

plot.pvals = readRDS("AOV_varimp_plot.rds")

>>>>

ANOVA - Anaylis of Variance. A statistical test that can be applied to find whether there is a significant difference between the means of more than two groups.
B2B - Business to Business 
CBR - Case based reasoning. A method of estimating by using the results of similar cases.
CRM - Customer Relationship Management. Refers to a common type of software used by businesses to record client and project details as well as employee timesheet records. Generally accessable to each employee in a company.
DAG - directed acyclic graph. A graph consisting of vertices connected by edges, where each edge directs a vertex to another vertex. The arrangement of the edges is such that a vertex cannot follow a path that loops back to itself.
IT - Information technology
BIM - Building Information Modelling. A digital representation of physical characteristics of an object or structure.

@@

Advantages of CBR include algorithmic efficiency as well as user engagement. In @Kim2004's study, it was noted that CBR models were simple to update with new data in comparison to neural networks, which must be slowly re-trained and re-tuned. New cases also did not need to have every variable completed which can occur often in real, messy data. Neural networks on the other hand require complete data. Another significant advantage of CBR is the ability for users to review chosen similar cases and make sense of the predicted cost, as opposed to neural networks which are a black box [@Kim2004][@Elfaki2014][@Kumar2007]. One disadvantage is that accuracy can be highly dependent on the number of selected cases [@Elfaki2014]. Overall, CBR may provide prediction accuracies between regression and neural networks but the results can be justified by the user - a valuable asset.

CBR can be described as a systematic method of expert judgment, where the decision maker manually compares similar projects from his or her experience [@Shepperd1996]. CBR that is performed manually through personal experience, without the assistance of algorithms, was the second most popular method for construction estimation in @Akintoye2000's survey in the UK. This indicates it is quite an intuitive method that allows decision makers to cross check their detailed cost estimates, although an algorithmic version of CBR has not found success in industry.

@@

In summary, research into cost estimation in the construction industry has trialled a variety of methods ranging from detailed manual analysis, algorithmic models, and case based reasoning. Neural networks generally outperformed all other systematic methods, such as regression and case based reasoning, but had the significant disadvantage of being a black box. Alternative methods such as boosted trees have the potential to perform as well as neural networks and provide insight into the structure of the algorithm. This is true for CBR as well, although it has not been shown to predict as accurately as neural networks. There exists a gap in the research in ways to intelligently combine CBR with machine learning methods that could predict well, provide insight to the structure, and engage the user in reviewing similar projects. Also, as stated previously, there exists a gap in the application of cost estimating in the construction industry to smaller components of the project, such as consulting and contracting companies who each face a similar problem in estimating their fixed fee. These companies could mobilise information in their internal project databases, as opposed to the current body of literature that used data from dozens of companies.

@@

Like the construction industry, the software industry faces the challenging task of estimating the cost of a project before formal engagement by their client. However, unlike construction projects, the major cost associated with the project is *effort* as opposed to the cost of building materials. A parallel can be drawn between software development effort estimation and consultants' effort estimation in the construction industry - which was highlighted as a gap in the research of the previous section.

Effort estimation is indeed a problem, and was demonstrated by a review of 6 surveys of IT projects between 1984 and 1994 by @Molokken2003. It revealed that 60-80% of IT projects encountered effort and/or schedule overruns, where the average overrun was by 30 to 40%. A survey of software managers asked whether they saw effort estimation as a problem, and 91% responded 'yes', while only 9% answered 'no' [@Moores1992]. The most popular response for the supposed reason for cost overruns was over-optimistic estimates (51%), which falls squarely within the theory of optimism bias, or the 'inside view' [@Flyvbjerg2011]. Given this widespread problem in the industry, plenty of research has been dedicated to the problem of effort prediction, with methods that can be categorised similarly to the construction industry: detailsed analysis, algorithmic modelling, and case based reasoning.

@@

Similar to the construction industry, expert judgment or detailed analysis, is the most widely practised method for effort estimation [@Shepperd1996][@Molokken2003]. This is despite many years of research being dedicated to developing algorithmic models and industry tools that, in the research context, outperform expert judgment. A survey by @Moores1992 found that software tools assisting a software project were in significant use, but only the project planning tools, not the cost estimating tools. This suggests a mental link for project managers to price jobs based on *details* of the software project as opposed to taking an outside view. Another study by @Heemstra1992 found that there was no evidence estimation accuracy improved when estimation tools were used. @Bergeron1992 similarly found jobs that used algorithmic models were actually associated with less accurate estimtes. However, this trend may be coincidental due to a lack of cases where estimation tools were used [@Molokken2003]. The scarcity of real project evidence that algorithmic estimation tools improve estimations and an unintuitive mental jump could be reasons contributing to detailed expert analysis remaining the most widespread technique for effort estimation. This trend mirrors what was found in the construction industry.

@@

Algorithmic methods to predict effort for IT project costs has been a popular subject of research. This section will assess the details and performance of the algorithmic models, important variables, applications and finally gaps in the literature.

Multiple studies have shown that neural networks definitively outperform regression models in effort estimation, although regression is the most popular method in the literature [@Finnie1997][@Pai2013][@Matson1993]. Interestingly, several studies have found that even if 15 or so variables are included, often only one variable contributes significantly to the models accuracy: size [@Shepperd1996][@Finnie1997][@Pai2013]. Size can refer to the expected number of lines of code in the software package or function point. Function point refers to the amount of business functionality expected from an information system product. This is easier to guess correctly at the beginning of a project than lines of code, and therefore leads to more accurate predictions [@Finnie1997]. One study used a single variable approach (size) with a linear coefficient and an exponential coefficient as follows:

$$
\begin{aligned}
 effort &= \alpha*size^{\beta} \\
 \alpha &= productivity \ coefficient \\
 \beta &= economies \ of\ scale\ coefficient \\
 size &= estimated \ lines \ of \ code
\end{aligned}
$$

This model was compared to CBR, which outperformed the model [@Shepperd1996].

@@

Algorithmic estimation tools have not been successfully adopted by the IT industry. This may be because no model has proved to be outstandingly successful at consistently predicting required effort [@Finnie1997]. Also, neural networks (the most accurate algorithm) lack the ability to capably explain its results and instill trust in decision makers which is another parallel to research from the construction industry [@Finnie1997].

As alluded to previosuly, studies on effort estimation as opposed to project cost estimation are concentrated in the software industry. This leaves a gap in studies for effort estimation in other industries such as the construction industry. Infrastructure projects do not have a function point variable and would need to utilise other information such as expected total project cost (which is notoriously hard to predict based on the previous section), client characteristics, and other project characteristics.

@@

In the IT industry, CBR applications have been researched using similar methodologies to the construction industry. The procedure is described as algorithmically finding similar projects using measurements of euclidean distance in n-dimensional space, where each dimension corresponds to a variable [@Shepperd1996]. Then, the similar cases are used to predict effort for the new case. In @Shepperd1996's study, linear regression was employed.

The literature found CBR to perform approximately equally to algorithmic methods such as neural networks which contrasts findings in the construction industry [@Shepperd1996][@Finnie1997]. It was also highlighted that CBR is intuitively similar to how an expert thinks about pricing projects and provides insight into the prediction that neural networks do not [@Finnie1997].

@@

Research into effort estimation in the software industry has followed a similar path to cost prediction studies on construction projects. Expert judgment via detailed analysis is still the predominant method for effort/cost estimation. Case based reasoning and algorithmic methods such as neural networks have found some success in effort estimation (at similar levels) but have not been successfully adopted by industry. It was noted that case based reasoning resonated more with decision makers because its predcitions could be better justified by the model. The visibility of model structure is very important for future work if the results are to be translated to industry. Effort estimation remains a problem, however agile methods of project management and effort management are changing the contractual approach of consulting in the software industry [@Badenfelt2011]. This innovative method keeps delivery times short (and therefore more predictable effort) and defines projects as something with a fixed cost and time. Therefore, if a change occurs, other less important elements must be discarded to maintain the fixed amount. Software development is a relatively new field in comparison to the construction industry and was able to quickly adopt this new model. In contrast, effort estimation in the construction industry faces a long tradition of fee structures that is difficult to reinvent. This leaves a gap in research of algorithimic models that predict effort (and not construction cost) for consulting companies in the construction industry. From findings in previous literature, the models should reveal their predictive structure to decision makers and be based on internal company data. Review of the literature as a whole reveals a lack in analysis of overall profit improvement a model could provide to a business. This could pesuade industry adoption and provide a framework around how models should affect decisions.

## Statistical and Machine Learning Methods

@@

AS stated, the aim of this project is to use statistical and machine learning techniques to model the profitability of projects. Therefore, in order to optimise the model, several statistical and machine learning techniques will be tested and it is prudent to review the range of algorithms that have been successfully applied in the literature. These vary from simple methods such as Linear Regression to complex, deep learning Neural Networks. The previous section highlighted the use of linear regression, neural networks, SVMs and in one case boosted trees for prediction, however research on other business problems utilised a wider range of methods. These include Naiive Bayes, Random Forests, and machine learned Bayesian Networks. The following section presents each method's advantages and disadvatanges and an example of a past successful application.

@@

Linear regression is one of the simplest and most popular statistical prediction methods [@Hastie2009]. An equation for a line is learned, which is defined by a constant and a coefficient times the value of each variable. The coefficients are chosen to minimise the residual sum of squares (least squared) [@Hastie2009]. The reason this method is popular is that it is simple, provides stable predictions, and can be adapted for categorical variables [@Seng2010]. The integration of categorical variables is called Anaylis of Variance (ANOVA) linear regression, and the algorithm compares group mean variances within the categorical variables.

There are several drawbacks to linear regression. For example assumptions must be made about the structure of the data. Predictor variables are assumed to be linearly related to the response variable and the variables to be normally distributed. Many real world phenomena do not correspond to these assumptions which can make it difficult, sometimes impossible, to produce useful results [@Hastie2009][@Putler2012]. Furthermore, response variables must be continous numeric variables and if two or more predictor variables are correlated, the values of their coefficients are highly unstable. On the other hand, correlated variables do not impact the predictive accuracy of the model. A final drawback is that the variable coefficients cannot be used to compare variable importance unless the variables have all been standardised to one another [@Putler2012]. This limits the amount of insight that can be drawn from a linear regression model.

Despite limitations of linear regression models, they have widespread business applications due to their simplicity and ease of application. Regression is used to predict risk in the finance and insurance industries, predict who to target in marketing exercises, and predict consumption spending in the field of economics [@Harrell2013]. Its simplicity and widespread application make linear regression a good starting point for analysis and can serve as a performance benchmark against complex models.

@@

The method behind linear regression can be adapted to predict a binary response variable (1 or 0). A linear equation is fit to the log of the odds of the binary response variable against each of the explanatory variables. The response variable (log odds) in the linear equation can then be translated back to odds by taking the inverse of log, the exponential. Finally the odds is converted to probability by the following equation:

$probability = \frac{odds}{1 + odds}$
[@Lowry2016]

This translates the original log odds response variable to a probability S curve bounded by 0 and 1. An example of this translation is graphically shown below:

![](images/lit_review/logreg.JPG)
[@Lowry2016]

The result of this linear fit and logistic translation is a model that predicts the probability (between 0 and 1) that an input case will result in a 0 or 1 (i.e. success or failure) [@Moore1989]. Similar pros and cons exist for logistic regression as linear regression, and it can also be used as a benchmark to compare other binary predictive models due to its speed and simplicity. Many business problems have binary response variables, such as yes/no, male/female, buy/do not buy, success/failure, or survival/death [@Moore1989].

@@

The naive bayes method works by making conditional independence assumptions about the explanatory variables in order to greatly simplify probability calculations for the response variable (the response variable must be categorical). More specifically, it is assumed that each variable is conditionally independent given the respose class, without considering other predictor variables. Mathematically this means that each variable's probability contribution can be simply calculated by looking at the probability of a response class given a single predictor value. A probability of each response class can be calculated for each explanatory variable. These are then combined for each possible response class using Bayes theorem, and the response class with the highest probability is chosen [@Provost2013].

The advantages of this method are that the conditional independence assumption enables very fast calculations and predictions. The method can perform very well for real world tasks because the assumption of independence does not damage predictions significantly. This is because if multiple variables are related, the variables independently still direct the prediction in the correct direction. The correlated variables will double or triple their emphasis on the predicted response variable, however this does not necessarily impact classification accuracy as the final step is to simply choose the class with the *highest* probability [@Provost2013]. This is fine for ranking. On the other hand, the output probabilities are not realisitic, and the values themselves should not be used [@Caruana2006]. Another disadvantage is that in order to calculate the class probability from a numeric variable, a distribution must be assumed (often gaussian) which the numeric data may not fit into neatly [@Caruana2006]. Finally, Naive Bayes classifers are known to not perform well in binary classification [@Caruana2006].

A real world success story for Naive Bayes classifiers hinged on their ability to be refined incrementally with each piece of new data, as opposed to re-calibrating the entire model. Naive Bayes was used in a complex spam detection system where new spam emails or toxic text themes could be quickly added to the filtering model. The Naive Bayes method provides a good benchmark to compare against more complex models that should outperform it [@Caruana2006].

@@

Decision trees are one of the simplest and most intuitive machine learning methods. There are several brands of basic decision tree algorithms including ID3, C4.5, CART, and CHAID with CART and C4.5 being the most popular [@Kabra2001]. Ravi Kumar and Ravi [@Kumar2007] recommended the CART algorithm as it is capable of solving both classification and regression problems whereas the remaining decision trees solve classification problems only. Single decision trees tend to overfit the data and provide low predictive accuracy but have favourable characteristics such as creating intuitive rules that a decision maker can follow in real-life scenarios. In addition, their results do not depend heavily on the skill of the data analyst [@Kabra2001][@Putler2012].

To combat the problems of low predictive accuracy and instability, ensemble tree methods were pioneered in the 1990's with success [@Breiman1996]. Three examples of these ensemble decision tree methods are boostrap aggregatiing (bagging), boosting, and random forests. Bagging creates multiple trees by sampling a different training data set (*with* replacement) for each tree and then combining tree results [@Breiman1996]. Random forests is an advanced form of bagging and builds on the idea of training multiple trees from the same data by sampling bootstrapped training sets with replacement. However, when creating each tree, a random subset of attributes (variables) are considered at each split. The reduced subset of attributes is resampled for each split in the tree. This allows dominant variables to be supressed for a fraction of the splits, allowing the algorithm to explore signals in weaker variables. It also prevents the trees from becoming too correlated [@Breiman2001]. The boosted decision tree approach applies gradient descent theory to a series of decision trees. The trees are limited to a certain depth to maintain simplicity, and each tree models the residuals (or errors) of the preceding tree. The limited depth of each tree prevents overfitting at each stage and the combined result of up to thousands of trees is very powerful [@Elith2008].

Ensemble decision trees have a different foundation to traditional statistical prediction methods such as regression. For this reason, there are some advantages and disadvantages. Decision trees use a progression of binary splits which means the data is first split by one variable at a certain value, then these subsets are split by another variable and so on in order to partition the response variable values into groups with similar values. The aim of each split is to achieve the most 'pure' split of the response values in a greedy fashion. If this process is visualised graphically, it is clear that non-linear relationships can be captured.

![](images/lit_review/dec_tree.JPG)
[@Jeevan2015]

In contrast to linear regression, the predictor variables do not need to be transformed as no assumptions are made about the data's statistical distributions [@Louppe2014][@Radenkovic2010]. Other advantages are that the ensemble methods are very fast in comparison to SVM's and neural networks but perform just as well and provide insights such as variable importance and variable relationships [@Sealfon2012]. Random forests in particular have the benefit being unaffected by noise in the data as well as the capacity to handle a large number of variables [@Sealfon2012]. Caruana [@Caruana2006] tested boosted trees, random forests, neural networks, SVM's, logistic regression and naive bayes on 11 binary classification problems and found that boosted trees performed best, followed by random forests. This demonstrates ensemble tree methods are capable of competing with high level machine learning algorithms.

Ensemble trees have numerous advantages however it is important to be aware of their limitations. Trees are not built on a probabilistic framework, and therefore their results can not be provided in this framework. For example, statistical confidence intervals for predictions are not available for standard ensemble methods [@Louppe2014]. This may or may not be required depending on the nature of the problem. Also, variable importance tables provided by random forests and boosted trees can be biased towards variables with many categories and another type of ensemble method, conditional forests, should be used [@Radenkovic2010]. The methods can also be prone to overfitting if not tuned carefully. Ensemble trees have many advantages particularly predictive power, but it is important to recognise that the best method for a problem often depends on the type of problem [@Louppe2014][@Caruana2006].

@@

A bayesian network is a graphical probabilistic model that illustrates the conditional dependencies between variables in a data set. The model is visually represented by a DAG (directed acyclic graph) and is capable of linking the conditional dependency between any variable to another variable or via other variables. The links are graphically represented by arcs or edges [@Heckerman1998]. This kind of conditional relationship is bayesian, where the probabilities in one node are conditional upon values in nodes directed to it as well as preceding nodes. Bayesian networks have found success in combining deterministic models with observed data as well as expert knowledge. The statistical relationships between variables can be determined from separate sources and manually combined into a single model [@Kragt2009].

Bayesian networks are excellent at representing knowledge from different origins, and then visually communicating results well to decision makers [@Kragt2009]. Therefore bayesian networks have become an excellent decision support tool. Network relationships can be machine learned, however this is not widely included as part of the suite of machine learning methods and is usually recommeneded for modelling complex networks. Drawbacks of bayesian networks include their inability to handle continuous numeric variables (all continuous variables must be discretised). This ensures all variable relationships are categorical [@Kragt2009]. Popular applications of bayesian networks are modelling uncertainty in natural resource management and modelling complex business network structures such as airports [@Wu2013].

@@

Neural networks were inspired by biological neural networks of the human nervous system and require a lot of training data relative to other methods. They are flexible in the range of relationships they can mimic and are not greatly dependent on the skills of the analyst (Putler & Krider, 2012). A disadvantage is that although the results are accurate, the algorithm is a black box - meaning the data is input and the results output without providing the user with insight into the methodology. Neural networks can perform high level tasks as a result of deep learning such as hand writing recognition, vehicle control, face recognition, and cancer detection [@Haykin2004].

The simplest type of SVM's calculate the linear boundary between two categories in feature space so that the category of future cases may be determined depending on which side of the boundary they fall. The location of the boundary is determined by the position of the widest clear margin that can be drawn between data points from the two categories, while allowing a certain number of misclassifications. The simple plot below demonstrates this concept and has no misclassifications:

![](images/lit_review/svm.JPG)
[@Provost2013]

This concept can be adapted to non linear boundaries by performing kernel transformations on prior variables as well as regression problems by introducing loss functions [@Auria2008]. Advantages of SVM's include the ability to model non linear relationships in data, and that no assumptions about the distributions of the data must be made as the model is non-probabilistic. They are also known to be robust over different samples and perform well in high-dimensional feature space [@Auria2008]. On the other hand, the output is in the form of distance to the boundary (as opposed to probability) and results are not transparent (similar to neural networks) [@Caruana2006]. Also, SVM's are known to be very slow to train and therefore often not suitable for industry purposes [@Auria2008]. Because SVM's perform well in high-dimensional space and can form non-linear rules they have been applied to protein classification (medicine) and text and image recognition [@Byun2002].

@@

Once

@@

plot.pvals = readRDS("report_data/AOV_varimp_plot.rds")
